Finally Tagless, Partially Evaluated

### 2008-08-14 tagless/tagless-final.tex
Various changes -- I'm in the middle of clarifying the self-interpretation section

%% TODO: add a ref 
%%  Lars Birkedal's master thesis
%%   http://www.itu.dk/people/birkedal/papers/paresm.ps.gz
%% and thank Eijiro Sumii

@@

Our main idea is to encode
\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
using compiler-generator (cogen) functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
\oleg{There is now IncopeDB.hs that defines at least R, L and C
 interpreters (no PE at present though) using DeBruijn indices. So we
 can say that (some of) our development is available in DeBruijn
 style. We won't talk about it in the paper though.}
In contrast to \S\ref{intro}, we encode binding using higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices. This makes the encoding convenient and
ensures that our object programs are closed.

@@

This interpreter~|RCN| is fully polymorphic over the answer type,
using higher-rank polymorphism through OCaml record types.
It could also be a functor parameterized over
the answer type.
\begin{code}
module RCN(W : sig type w end)= struct
  type ('c,'sv,'dv) repr = ('sv -> W.w) -> W.w
  ...
\end{code}
This allows a translation of these techniques to, for example, Standard ML.
Unfortunately, the loss of polymorphism in the answer type makes the 
CPS evaluator less pleasant to use than the other implementations.

@@

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the type of |lam f|. Whereas
|int| and |add| return the (abbreviated) type
|('c, ..., (int -> 'w) -> 'w) S.repr|,
the type of \texttt{lam (add (int~1))} is
\begin{code}
('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr
\end{code}
Hence, to write the type equation defining |CPST.repr| we again need
a type function with a typecase distinction, similar to |repr_pe|
in~\S\ref{S:PE-solution}. Alas, the type function we need is not
identical to |repr_pe|, so we need to add another type argument to
|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
terms in previous implementations of |Symantics| stay unchanged, but the
|repr| type equations in those implementations have to take a new
(phantom) type argument.

For brevity, we just
use the module returned by |CPST| as is. Because it does not
match the signature |Symantics|, we cannot apply the |EX| functor to it.
Nevertheless, we can write the tests.
\begin{code}
module T = struct
  module M = CPST(C)
  open M
  let test1 () = (* same as before *)
       app (lam (fun x -> x)) (bool true)
  let testpowfix () = ... (* same as before *)
  let testpowfix7 = (* same as before *)
       lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
We instantiate |CPST| with the desired base interpreter~|C|,
then use the result |M| to
interpret object terms. Those terms are \emph{exactly} as before.
Having to textually copy the terms is the
price we pay for this simplified treatment.
Our discussion of self\hyp interpretation in~\S\ref{selfinterp} shows
that this copying is not frivolous but represents plugging a term into
a context, which is one of the many faces of polymorphism.

@@

We can modify a CBN or CBV CPS transformation to pass a piece of state
along with the continuation. This technique lets us support mutable
state. As Figure~\ref{fig:state} shows, we extend our object language
with three imperative features.
\begin{enumerate}
    \item ``$\deref$'' gets the current state;
    \item ``$\set e$'' sets the state to the value of~$e$ and returns
        the previous value of the state;
    \item the let-form ``$\lapp{e_1}{x}e_2$'' evaluates $e_1$
        before~$e_2$ even if $e_2$ does not use~$x$.
\end{enumerate}
If $x$ does not appear in~$e_2$, then ``$\lapp{e_1}{x}e_2$'' is same as
the more familiar sequencing form ``$e_1;e_2$''.
We can embed this extended object language into OCaml by extending the
|Symantics| signature in Figure~\ref{fig:ocaml}.
\begin{code}
module type SymSI = sig
  include Symantics
  type state
  type 'c states       (* static version of the state *)
  val lapp : (('c,'sa,'da) repr as 'x) -> ('x -> 'y)
             -> (('c,'sb,'db) repr as 'y)
  val deref : unit -> ('c, 'c states, state) repr
  val set   : (('c, 'c states, state) repr as 'x) -> 'x
end
\end{code}
In HOAS\@, we write the term ``$\lapp{e_1}{x}e_2$'' as |lapp e1 (fun x -> e2)|;
the type of |lapp| is that of function application with
the two arguments swapped.  We can encode the term
``$\lapp{\deref}{x} (\set 2;\, x+\deref)$''
as the OCaml functor
\begin{code}
module EXSI_INT(S: SymSI
  with type state = int and type 'c states = int) = struct open S
  let test1 () = lapp (deref ()) (fun x -> 
                  lapp (set (int 2)) (fun _ -> add x (deref ())))
end
\end{code}
The accompanying source code shows several more tests, including
a test for higher-order state and a power function that uses state
as the accumulator.
Because the |SymSI| signature extends |Symantics|, any encoding of
a term in the pure object language (that is, any functor that takes
a |Symantics| module as argument) can also be used as a term in the
extended object language (for example, applied to an implementation of
|SymSI|).

@@

|RCPSI| matches the |Symantics| signature and implements the
unextended object language: we can pass |RCPSI|
to the functor |EX| (Fig.~\ref{fig:ocaml-simple}) and run 
the example |test1| from there. The main use for |RCPSI| is to interpret the 
extended object language.
\begin{code}
module EXPSI_INT = EXSI_INT(RCPSI)
let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100
val cpsitesti1 : int = 102
\end{code}

It is worthwhile reiterating how close this implementation is to the CPS
interpreter, and that adding state and imperative features needed no new
techniques.
We can also add mutable references to the object language using
mutable references of the metalanguage, as shown in the accompanying
code.
\fi
\fi

@@

Given an \emph{encoding} of each object term~$e$ as an \emph{object} term~$\Encode{e}$,
a \emph{self\hyp interpreter} is usually defined as an object
function~$\si$ such that any object term~$e$ is observationally
equivalent to the object application $\si\Encode{e}$
\citep{jones-partial,taha-tag,Danvy-tagging-encoding}.
A partial evaluator~$\pe$ maps object terms~$e$ to observationally
equivalent object terms~$\pe(e)$.  It is said to be
\emph{optimal} with respect to~|si| \citep{jones-challenging} if $\pe(\si\Encode{e})$
is equal to~$e$ (up to $\alpha$\hyp conversion, or in some accounts, no
less efficient than~$e$).

Intuitively, self\hyp interpretation is straightforward in our
framework: the functions comprising the interpreters
in~\S\ref{S:interpreter-RL} may just as well be written in our object
language:~Fig.~\ref{fig:self-eval}.
\begin{figure}
\begin{align*}
    \ident{int} &= \fun{x} x \\
    \ident{add} &= \fun{x} \fun{y} x+y \\
    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}
        \displaybreak[0] \\
    \ident{lam} &= \fun{f} f \\
    \ident{app} &= \fun{f} \fun{x} fx \\
    \ident{fix} &= \fun{g} \fix{f} \fun{x} gfx
\end{align*}
\caption{The \emph{object} functions implementing an
evaluator.  (We use the number~$0$ in lieu of a unit value.)}
\label{fig:self-eval}
\end{figure}
We thus map each object term~$e$ to an object term~$\encode{e}$ as follows.
We call this mapping \emph{pre-encoding}.
\begin{align*}
    \encode{x} &= x \\
    \encode{n} &= \ident{int}\, n \\
    \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} \\
    \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$}
        \displaybreak[0] \\
    \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
    \encode{fx} &= \ident{app} \encode{f} \encode{x} \\
    \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e})
\end{align*}
The metavariables $x$ and~$n$ stand for a variable and an integer,
respectively.
This pre-encoding is just like how we represent object terms in the
metalanguage in the preceding sections, but it produces
terms in the object language rather than the metalanguage.

To evaluate~$\encode{e}$, then, we
instantiate the free variables in~$\encode{e}$ such as $\ident{int}$,
$\ident{lam}$, and $\ident{add}$ by their definitions given in
Fig.~\ref{fig:self-eval}.  For
example, the familiar object term $(\fun{x}x)\True$ pre-encodes to
\begin{equation*}
    \encode{(\fun{x}x)\True} = \ident{app}
    (\ident{lam} (\fun{x} x))\, (\ident{bool} \True),
\end{equation*}
and to evaluate this pre-encoded term is to evaluate the object term
\begin{equation*}
    (\fun{f} \fun{x} fx) \,
    ((\fun{f} f) (\fun{x} x))\, ((\fun{b} b) \True).
\end{equation*}
Because the evaluator in Fig.~\ref{fig:self-eval} mostly consists 
of glorified identity
functions, our simple partial evaluator reduces the
result of this instantiation to~$e$.  In general, to
interpret~$\encode{e}$ using an interpreter is to instantiate its free
variables by that interpreter's definitions.

@@

Any approach to self\hyp interpretation needs to spell out first how to
encode object terms~$e$ to object terms~$\Encode{e}$, and then how to
interpret~$\Encode{e}$ in the object language.  For our approach, we
want to define encoding in terms of pre-encoding, and interpretation
using some notion of instantiation.  Unfortunately, the simple type
structure of our object language hinders both tasks.  To continue with
the example term above, we could try to define
\begin{equation*}
    \Encode{e} =
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}} \encode{e},
\end{equation*}
in particular
\begin{equation*}
    \Encode{(\fun{x}x)\True} =
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True).
\end{equation*}
To type-check this encoded term, we give the variable $\ident{lam}$ the
simple type $(\BB\to\BB)\to\BB\to\BB$.
We then define the self\hyp interpreter
\begin{equation*}
    \si = \fun{e} e
    (\fun{f} \fun{x} fx)
    (\fun{f} f)
    (\fun{b} b)
\end{equation*}
and apply it to the encoded term.  The result is the object term
\begin{equation*}
    \bigl(\fun{e} e (\fun{f} \fun{x} fx) (\fun{f} f) (\fun{b} b)\bigr)
    \bigl(
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True)
    \bigr),
\end{equation*}
which partially evaluates to~$\True$ easily.
However, encoding fails on a term with multiple $\lambda$\hyp
abstractions at different types.  For example, the pre-encoding
\begin{equation*}
    \encode{\fun{f}\fun{x}fx}
    = \ident{lam} (\fun{f} \ident{lam} (\fun{x} \ident{app} f x))
\end{equation*}
does not type-check in any typing environment, because $\ident{lam}$ needs
to take two incompatible types.  In sum, we need more polymorphism in the
object type system to type $\ident{lam}$, $\ident{app}$, $\ident{fix}$,
and~$\ident{if\_}$ (and $\Encode{e}$ and~$\si$).
(The polytypes in |Symantics| given by Haskell's type classes and OCaml's
modules supply this polymorphism.)  Moreover, we need to encode any
polymorphism of the object language \emph{into} the object language to achieve
self\hyp interpretation.

@@

Instead of adding higher-rank and higher-kind polymorphism to our object
language (along with polymorphism over kinds!), we add let-bound polymorphism.
As usual,
we can add a new typing rule
\begin{equation*}
    \let\vcenter\vbox
    \begin{prooftree}
        e_1:t_1 \quad \subst{e_2}{x}{e_1}:t_2
        \justifies \be{x=e_1} e_2 : t_2
    \end{prooftree}
    .
\end{equation*}
The pre-encoding of a let\hyp expression is trivial.
\begin{align*}
    \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
\intertext{A \emph{context} is an object term with a hole~$[~]$.  The
hole may occur under a binder, so plugging a term into the context may
capture free variables of the term.  By pre-encoding a hole to a hole,
we extend pre-encoding from a translation on terms to one on
contexts.}
    \encode{[~]} \quad &= \quad [~]
\end{align*}
We define an interpreter in the object language to be not a term but
a context.  For example, the evaluator is the context
\begin{align*}
    \si[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} x&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y&} \\
        \be{\ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}&} \\
        \be{\ident{lam} &= \fun{f} f&} \\
        \be{\ident{app} &= \fun{f} \fun{x} fx&} \\
        \be{\ident{fix} &= \fun{g} \fix{f} \fun{x} gfx&} [~],
    \end{tabular}
\intertext{and the size\hyp measurer at the end of~\S\ref{S:interpreter-RL} is the context}
    \mathrm{SZ}[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} 1&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y+1&} \\
        \be{\ident{if\_}&= b + t\,0 + e\,0&} \\
        \be{\ident{lam} &= \fun{f} f\,0 + 1&} \\
        \be{\ident{app} &= \fun{f} \fun{x} f + x + 1&} \\
        \be{\ident{fix} &= \fun{g} g\,0 + 1&} [~].
    \end{tabular}
\end{align*}
To interpret an object term~$e$ using an interpreter $I[~]$ is to
evaluate the object term~$I[\encode{e}]$.  $\si$~is
a self\hyp interpreter in the following sense.
\begin{proposition}
    $\si[\encode{e}]$ is observationally equivalent to~$e$.
\end{proposition}
As a corollary, we can pre-encode $\si$ itself as
a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
$\si$ can interpret itself.  Our partial evaluator is optimal with respect to
the self\hyp interpreter~$\si$.
\begin{proposition}
    Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
    Then the object terms $\pe(\si[\encode{e}])$ and~$\pe(e)$ are
    either both undefined or both defined and
    equal up to $\alpha$\hyp conversion.
\end{proposition}

@@

We always type-check a pre-encoded term~$\encode{e}$ and its interpreter~$I[~]$
together, never separately.  This treatment has the drawback that
we must duplicate a pre-encoded term in order to interpret it in
multiple ways.  The meta-notions of contexts and plugging may seem ad
hoc, but in fact they just reflect the type-class and module
machinery that we have been using in this paper all along.

@@

It is common to implement an embedded DSL by providing multiple
interpretations of host\hyp language pervasives such as addition and
application.  It is also common to use phantom types to rule out
ill-typed object terms, as done in Lava
\citep{Lava} and by \citet{Rhiger-thesis}. However, these approaches
are not tagless because they still use universal types, such as Lava's
\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Fig.~2.2)
and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
pattern matching.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\citep{Rhiger-thesis}, whereas our proofs are trivial.
Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1)
\end{comment}

>>>>

%% TODO: add a ref 
%%  Lars Birkedal's master thesis
%%   http://www.itu.dk/people/birkedal/papers/paresm.ps.gz

@@

Our main idea is to encode
\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
using combinator functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
\begin{comment}
\oleg{There is now IncopeDB.hs that defines at least R, L and C
 interpreters (no PE at present though) using de Bruijn indices. So we
 can say that (some of) our development is available in de Bruijn
 style. We won't talk about it in the paper though.}
\end{comment}
In contrast to \S\ref{intro}, we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensures that our object programs are closed.
We find HOAS to be more convenient, though we could have continued to
use de Bruijn indices.

@@

This interpreter~|RCN| is fully polymorphic over the answer type,
using higher-rank polymorphism through OCaml record types.
To avoid this higher-rank polymorphism in the core language, which
Standard ML does not support, we could also define |RCN| as
a functor parameterized over
the answer type.
\begin{code}
module RCN(W : sig type w end)= struct
  type ('c,'sv,'dv) repr = ('sv -> W.w) -> W.w
  ...
\end{code}
This allows a translation of these techniques to, for example, Standard ML\@.
Unfortunately, the loss of polymorphism in the answer type makes the 
CPS evaluator less pleasant to use than the other implementations.

@@

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the type of |lam f|. Whereas
|int| and |add| return the (abbreviated) type
|('c, ..., (int -> 'w) -> 'w) S.repr|,
the type of \texttt{lam (add (int~1))} is
\begin{code}
('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr
\end{code}
Hence, to write the type equation defining |CPST.repr| we again need
a type function with a typecase distinction, similar to |repr_pe|
in~\S\ref{S:PE-solution}. Alas, the type function we need is not
identical to |repr_pe|, so we need to add another type argument to
|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
terms in previous implementations of |Symantics| stay unchanged, but the
|repr| type equations in those implementations have to take a new
(phantom) type argument.
The verbosity of these types is the only difficulty in defining a
new signature to replace |Symantics| that the output of |CPST| matches.

For brevity, we just
use the module returned by |CPST| as is. Because it does not
match the signature |Symantics|, we cannot apply the |EX| functor to it.
Nevertheless, we can write the tests.
\begin{code}
module T = struct
  module M = CPST(C)
  open M
  let test1 () =
       app (lam (fun x -> x)) (bool true)  (* same as before *)
  let testpowfix () = ...                  (* same as before *)
  let testpowfix7 =                        (* same as before *)
       lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
We instantiate |CPST| with the desired base interpreter~|C|,
then use the result |M| to
interpret object terms. Those terms are \emph{exactly} as before.
Having to textually copy the terms is the
price we pay for this simplified treatment.
Our discussion of self\hyp interpretation in~\S\ref{selfinterp} shows
that this copying is not frivolous but represents plugging a term into
a context, which is one of the many faces of polymorphism.

@@

We can modify a CBN or CBV CPS transformation to pass a piece of state
along with the continuation. This technique lets us support mutable
state. As Figure~\ref{fig:state} shows, we extend our object language
with three imperative features.
\begin{enumerate}
    \item ``$\deref$'' gets the current state;
    \item ``$\set e$'' sets the state to the value of~$e$ and returns
        the previous value of the state;
    \item the let-form ``$\lapp{e_1}{x}e_2$'' evaluates $e_1$
        before~$e_2$ even if $e_2$ does not use~$x$.
\end{enumerate}
If $x$ does not appear in~$e_2$, then ``$\lapp{e_1}{x}e_2$'' is same as
the more familiar sequencing form ``$e_1;e_2$''.
We can embed this extended object language into OCaml by extending the
|Symantics| signature in Figure~\ref{fig:ocaml}.
\begin{code}
module type SymSI = sig
  include Symantics
  type state
  type 'c states       (* static version of the state *)
  val lapp : (('c,'sa,'da) repr as 'x) -> ('x -> 'y)
             -> (('c,'sb,'db) repr as 'y)
  val deref : unit -> ('c, 'c states, state) repr
  val set   : (('c, 'c states, state) repr as 'x) -> 'x
end
\end{code}
In HOAS\@, we write the term ``$\lapp{e_1}{x}e_2$'' as |lapp e1 (fun x -> e2)|;
the type of |lapp| is that of function application with
the two arguments swapped.  We can encode the term
``$\lapp{\deref}{x} (\set 2;\, x+\deref)$''
as the OCaml functor
\begin{code}
module EXSI_INT(S: SymSI
  with type state = int and type 'c states = int) = struct open S
  let test1 () = lapp (deref ()) (fun x -> 
                  lapp (set (int 2)) (fun _ -> add x (deref ())))
end
\end{code}
The accompanying source code shows several more tests, including
a test for higher-order state and a power function that uses state
as the accumulator.

@@

Because the |SymSI| signature extends |Symantics|, any encoding of
a term in the pure object language (that is, any functor that takes
a |Symantics| module as argument) can also be used as a term in the
extended object language (for example, applied to an implementation of
|SymSI|).  In particular,
|RCPSI| matches the |Symantics| signature and implements the
unextended object language: we can pass |RCPSI|
to the functor |EX| (Fig.~\ref{fig:ocaml-simple}) and run 
the example |test1| from there. The main use for |RCPSI| is to interpret the 
extended object language.
\begin{code}
module EXPSI_INT = EXSI_INT(RCPSI)
let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100
val cpsitesti1 : int = 102
\end{code}

We reiterate that this implementation adding state and imperative
features is very close to the CPS interpreter and uses no new
techniques.
We can also add mutable references to the object language using
mutable references of the metalanguage, as shown in the accompanying
code.
\fi
\fi

@@

Given an \emph{encoding} of each object term~$e$ as an \emph{object} term~$\Encode{e}$,
a \emph{self\hyp interpreter} is usually defined as an object
function~$\si$ such that any object term~$e$ is observationally
equivalent to the object application $\si\Encode{e}$
\citep{jones-partial,taha-tag,Danvy-tagging-encoding}.
A partial evaluator~$\pe$ maps object terms~$e$ to observationally
equivalent object terms~$\pe(e)$.  It is said to be
\emph{optimal} with respect to~|si| \citep{jones-challenging} if $\pe(\si\Encode{e})$
is $\alpha$\hyp equivalent to~$e$ (or in some accounts, no
less efficient than~$e$).

Self\hyp interpretation in our framework is straightforward at the
term level: the functions comprising the interpreters
in~\S\ref{S:interpreter-RL} may as well be written in our object
language, as in Fig.~\ref{fig:self-eval}.
\begin{figure}
\begin{align*}
    \ident{int} &= \fun{x} x \\
    \ident{add} &= \fun{x} \fun{y} x+y \\
    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}
        \displaybreak[0] \\
    \ident{lam} &= \fun{f} f \\
    \ident{app} &= \fun{f} \fun{x} fx \\
    \ident{fix} &= \fun{g} \fix{f} \fun{x} gfx
\end{align*}
\caption{The \emph{object} functions implementing an
evaluator.  (We use the number~$0$ in lieu of a unit value.)}
\label{fig:self-eval}
\end{figure}
We thus map each object term~$e$ to an object term~$\encode{e}$ as follows.
We call this mapping \emph{pre-encoding}.
\begin{equation}
\begin{split}
    \encode{x} &= x \\
    \encode{n} &= \ident{int}\, n \\
    \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
    \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} \\
    \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$} \\
    \encode{fx} &= \ident{app} \encode{f} \encode{x} \\
    \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e})
\end{split}
\end{equation}
The metavariables $x$ and~$n$ stand for a variable and an integer,
respectively.
This pre-encoding is just like how we represent object terms in the
metalanguage in the preceding sections, but it produces
terms in the object language rather than the metalanguage.

To evaluate~$\encode{e}$, then, we
instantiate the free variables in~$\encode{e}$ such as $\ident{int}$,
$\ident{lam}$, and $\ident{add}$ by their definitions given in
Fig.~\ref{fig:self-eval}.  For
example, the familiar object term $(\fun{x}x)\True$ pre-encodes to
\begin{equation}
    \encode{(\fun{x}x)\True} = \ident{app}
    (\ident{lam} (\fun{x} x))\, (\ident{bool}\, \True),
\end{equation}
and to evaluate this pre-encoded term is to evaluate the object term
\begin{equation}
    (\fun{f} \fun{x} fx) \,
    ((\fun{f} f) (\fun{x} x))\, ((\fun{b} b) \True).
\end{equation}
Because the evaluator in Fig.~\ref{fig:self-eval} mostly consists 
of glorified identity
functions, our simple partial evaluator reduces the
result of this instantiation to~$e$.  In general, to
interpret~$\encode{e}$ using an interpreter is to instantiate its free
variables by that interpreter's definitions.

@@

Any approach to self\hyp interpretation needs to spell out first how to
encode object terms~$e$ to object terms~$\Encode{e}$, and then how to
interpret~$\Encode{e}$ in the object language.  For our approach, we
want to define encoding in terms of pre-encoding, and interpretation
using some notion of instantiation.  Unfortunately, the simple type
structure of our object language hinders both tasks.  To continue with
the example term above, we could try to define
\begin{equation}
    \Encode{e} =
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}} \encode{e},
\end{equation}
in particular
\begin{equation}
    \Encode{(\fun{x}x)\True} =
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool}\, \True).
\end{equation}
To type-check this encoded term, we give the bound variable $\ident{lam}$ the
simple type $(\BB\to\BB)\to\BB\to\BB$.
We then define the self\hyp interpreter
\begin{equation}
    \si = \fun{e} e
    (\fun{f} \fun{x} fx)
    (\fun{f} f)
    (\fun{b} b)
\end{equation}
and apply it to the encoded term.  The result is the object term
\begin{equation}
\kern-\mintagsep
    \bigl(\fun{e} e (\fun{f} \fun{x} fx) (\fun{f} f) (\fun{b} b)\bigr)
    \bigl(
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool}\, \True)
    \bigr),
\end{equation}
which partially evaluates to~$\True$ easily.
However, encoding fails on a term with multiple $\lambda$\hyp
abstractions at different types.  For example, the pre-encoding
\begin{equation}
\label{e:pre-encoding-example}
    \encode{\fun{f}\fun{x}fx}
    = \ident{lam} (\fun{f} \ident{lam} (\fun{x} \ident{app} f x))
\end{equation}
does not type-check in any typing environment, because $\ident{lam}$ needs
to take two incompatible types.  In sum, we need more polymorphism in the
object type system to type $\ident{lam}$, $\ident{app}$, $\ident{fix}$,
and~$\ident{if\_}$ (and $\Encode{e}$ and~$\si$).
(The polytypes in |Symantics| given by Haskell's type classes and OCaml's
modules supply this polymorphism.)  Moreover, we need to encode any
polymorphism of the object language \emph{into} the object language to achieve
\emph{self}\hyp interpretation.

@@

Instead of adding higher-rank and higher-kind polymorphism to our object
language (along with polymorphism over kinds!), we add let-bound polymorphism.
As usual,
we can add a new typing rule
\begin{equation}
\label{e:let}
    \let\vcenter\vbox
    \begin{prooftree}
        e_1:t_1 \quad \subst{e_2}{x}{e_1}:t_2
        \justifies \be{x=e_1} e_2 : t_2
    \end{prooftree}
    .
\end{equation}
The pre-encoding of a let\hyp expression is trivial.
\begin{align}
    \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
\intertext{A \emph{context} is an object term with a hole~$[~]$.  The
hole may occur under a binder, so plugging a term into the context may
capture free variables of the term.  By pre-encoding a hole to a hole,
we extend pre-encoding from a translation on terms to one on
contexts.}
    \encode{[~]} \quad &= \quad [~]
\end{align}
We define an interpreter in the object language to be not a term but
a context.  For example, the evaluator is the context
\begin{align}
\label{e:evaluator}
    \si[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} x&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y&} \\
        \be{\ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}&} \\
        \be{\ident{lam} &= \fun{f} f&} \\
        \be{\ident{app} &= \fun{f} \fun{x} fx&} \\
        \be{\ident{fix} &= \fun{g} \fix{f} \fun{x} gfx&} [~],
    \end{tabular}
\intertext{and the size\hyp measurer at the end of~\S\ref{S:interpreter-RL} is the context}
\label{e:size-measurer}
    \mathrm{SZ}[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} 1&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y+1&} \\
        \be{\ident{if\_}&= b + t\,0 + e\,0&} \\
        \be{\ident{lam} &= \fun{f} f\,0 + 1&} \\
        \be{\ident{app} &= \fun{f} \fun{x} f + x + 1&} \\
        \be{\ident{fix} &= \fun{g} g\,0 + 1&} [~].
    \end{tabular}
\end{align}
To interpret an object term~$e$ using an interpreter $I[~]$ is to
evaluate the object term~$I[\encode{e}]$.  $\si$~is
a self\hyp interpreter in the following sense.
\begin{proposition}
    $\si[\encode{e}]$ is observationally equivalent to~$e$.
\end{proposition}
As a corollary, we can pre-encode $\si$ itself as
a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
$\si$ can interpret itself.  Our partial evaluator is optimal with respect to
the self\hyp interpreter~$\si$.
\begin{proposition}
    Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
    Then the object terms $\pe(\si[\encode{e}])$ and~$\pe(e)$ are
    either both undefined or both defined and
    equal up to $\alpha$\hyp conversion.
\end{proposition}

@@

We always type-check a pre-encoded term~$\encode{e}$ in the context of
a particular interpreter~$I[~]$, never alone.  For example, to
type-check the pre-encoded term~\eqref{e:pre-encoding-example}, we must
plug it into an interpreter, such as the evaluator~\eqref{e:evaluator}.
This treatment has the drawback that we must duplicate the pre-encoding
of a term in order to interpret it in multiple ways, that is, to plug it
into multiple interpreters such as the evaluator~\eqref{e:evaluator} and
the size\hyp measurer~\eqref{e:size-measurer}.  In return, we avoid
adding polymorphism to the object language's type system, because we can
state the let rule~\eqref{e:let} in terms of substitution rather than by
generalizing and instantiating types.  In other words, we use contexts
and plugging in the object language in place of the type-class and
module machinery in the metalanguages.

@@

Our separation between the |Symantics| interface and its many
implementations codifies the common practice of implementing an embedded
DSL by specifying an abstract syntax of object\hyp language pervasives,
such as addition and application, then providing multiple
interpretations of them.  \Citet{JonesNielsen-D58} prefigured this
separation when they decomposed a denotational definition of an untyped
object language into a core semantics (which we call abstract syntax)
and multiple interpretations.  When implementing an embedded DSL in this
way, it is also common practice to use phantom types to rule out
ill-typed object terms, as done in Lava
\citep{Lava} and by \citet{Rhiger-thesis}. However, these two approaches
are not tagless because they still use universal types, such as Lava's
\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Fig.~2.2)
and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
pattern matching.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\citep{Rhiger-thesis}, whereas our proofs are trivial.
Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1)
\end{comment}
